---
title: "Pathology prediction in rowers"
author: "Félix BOUDRY"
---

## Context

We aim at detecting pathological athletes during an altitude training camp. To do so we used urine sample analysed by NMR. Here we try different models to detect the pathological samples to determine which one is the best.

```{r}
#| label: config
#| include: false
#| fig-align: center
set.seed(42)
library(tidyverse)
library(data.table)
library(knitr)
library(janitor)
library(psych)
library(kableExtra)
library(caret)
library(pls)
library(vip)
library(factoextra)
library(cluster)
library(ggfortify)
library(magrittr)
library(plotly)
library(ggpubr)
source("functions.R")

theme_set(new = theme_bw())
```

# Data description

```{r}
#| label: data-description
antrop_data <- fread(input = "Data/Subjects.csv") |>
  clean_names() |>
  select(c("age", "height", "weight"))
describe(x = antrop_data, na.rm = T) |>
  round(digits = 0) |>
  select(c("mean", "sd", "median", "min", "max")) |>
  kable(
    caption = paste0("Anthropological values (n = ", nrow(x = antrop_data), ")"),
    table.attr = 'data-quarto-disable-processing="true"'
  ) |>
  kable_styling(
    bootstrap_options = c("striped"),
    full_width = FALSE
  )
```

# Pre-processing

Pre-processing was done using [npflow](https://nmrprocflow.org).

```{r}
#| label: pre-processing
data_pre_post <- fread(input = "Data/data_Patho_first_day.csv") |>
  column_to_rownames(var = "Samplecode") |>
  mutate_at(
    .vars = vars(-"Patho", -"Patho_num"),
    .funs = \(x) ReplaceMissingByLoD(int.mat = x)
  ) |>
  mutate_at(
    .vars = vars(-"Patho", -"Patho_num"),
    .funs = \(x) scale(x = x)
  ) |>
  as.data.frame()

test_pre_post <- fread(input = "Data/test_set_first.csv") |>
  column_to_rownames(var = "Samplecode") |>
  mutate_at(
    .vars = vars(-"Patho", -"Patho_num"),
    .funs = \(x) ReplaceMissingByLoD(int.mat = x)
  ) |>
  mutate_at(
    .vars = vars(-"Patho", -"Patho_num"),
    .funs = \(x) scale(x = x)
  ) |>
  as.data.frame()

data_rand <- fread(input = "Data/data_Patho_rand.csv") |>
  column_to_rownames(var = "Samplecode") |>
  mutate_at(
    .vars = vars(-"Patho", -"Patho_num"),
    .funs = \(x) ReplaceMissingByLoD(int.mat = x)
  ) |>
  mutate_at(
    .vars = vars(-"Patho", -"Patho_num"),
    .funs = \(x) scale(x = x)
  ) |>
  as.data.frame()

test_rand <- fread(input = "Data/test_set_rand.csv") |>
  column_to_rownames(var = "Samplecode") |>
  mutate_at(
    .vars = vars(-"Patho", -"Patho_num"),
    .funs = \(x) ReplaceMissingByLoD(int.mat = x)
  ) |>
  mutate_at(
    .vars = vars(-"Patho", -"Patho_num"),
    .funs = \(x) scale(x = x)
  ) |>
  as.data.frame()
```

# First vs. pathological comparison

We compare data from the first day with data from the day symptoms were observed. Do not forget that the subjects are training in altitude and an effect of chronic hypoxic exposure can be observed.

## k-means

```{r}
#| label: km-pp
#| fig-align: center
km_res <- kmeans(
  x = select(.data = data_pre_post, -c("Patho", "Patho_num")),
  centers = 2L,
  nstart = 25L
)
km_cluster <- km_res$cluster - 1
fviz_cluster(
  object = km_res,
  data = select(.data = data_pre_post, -c("Patho", "Patho_num")),
  geom = "none"
) +
  geom_point(
    mapping = aes(shape = data_pre_post$Patho),
    size = 2
  ) +
  labs(shape = "Status")
confusionMatrix(data = as.factor(x = km_cluster), as.factor(x = data_pre_post$Patho_num)) |>
  draw_confusion_matrix()
```

## Hierarchical clustering

```{r}
#| label: hc-pp
#| fig-align: center
hc_res <- agnes(x = select(.data = data_pre_post, -c("Patho", "Patho_num")), method = "ward")
hclust_plot(clust_data = hc_res, target = data_pre_post$Patho_num)
confusionMatrix(
  data = hc_res |> cutree(k = 2L) |> subtract(e1 = 1) |> as.factor(),
  data_pre_post$Patho_num |> as.factor()
) |> draw_confusion_matrix()
```

## PCA

```{r}
#| label: pca-pp
#| fig-align: center
pca_res <- prcomp(x = select(.data = data_pre_post, -c("Patho", "Patho_num")))
autoplot(
  object = pca_res,
  data = select(.data = data_pre_post, -c("Patho", "Patho_num")),
  loadings.label.size = 3
) +
  geom_point(
    mapping = aes(color = data_pre_post$Patho),
    size = 2
  ) +
  stat_ellipse(mapping = aes(group = data_pre_post$Patho, colour = data_pre_post$Patho)) +
  labs(color = "Patho")
```

## PLS-DA

```{r}
#| label: plsda-pp
#| fig-align: center
pls_res <- mixOmics::plsda(
  X = select(.data = data_pre_post, -c("Patho", "Patho_num")),
  Y = as.factor(x = data_pre_post$Patho),
  ncomp = 5
)
pc1_var <- pls_res$prop_expl_var$X[[1]] * 100
pc2_var <- pls_res$prop_expl_var$X[[2]] * 100
ggplot(
  data = pls_res$input.X,
  mapping = aes(x = pls_res$variates$X[, 1], y = pls_res$variates$X[, 2])
) +
  geom_point(
    mapping = aes(color = data_pre_post$Patho),
    size = 2
  ) +
  stat_ellipse(mapping = aes(group = data_pre_post$Patho, colour = data_pre_post$Patho)) +
  xlab(label = paste0("PC1 (", pc1_var |> round(digits = 1), "%)")) +
  ylab(label = paste0("PC2 (", pc2_var |> round(digits = 1), "%)")) +
  labs(color = "Patho")

pls_res <- caret::plsda(
  x = select(.data = data_pre_post, -c("Patho", "Patho_num")),
  as.factor(x = data_pre_post$Patho),
  ncomp = 5,
  validation = "CV"
)
my_vips <- vip(object = pls_res, metric = "rmse")
plot(x = my_vips)
plot_data <- data_pre_post |>
  select(c("Patho", my_vips$data$Variable)) |>
  pivot_longer(
    cols = starts_with(match = "B"),
    names_to = "Buckets",
    values_to = "Value"
  )
buckets_stats <- lapply(
  my_vips$data$Variable,
  \(x) wilcox.test(data_pre_post[data_pre_post$Patho == "Pathological", x], data_pre_post[data_pre_post$Patho == "Healthy", x])$p.value
) |>
  `names<-`(my_vips$data$Variable) |>
  lapply(p.adjust, method = "BH")
ggplot(
  data = plot_data,
  mapping = aes(x = Buckets, y = Value, fill = Patho)
) +
  geom_boxplot() +
  stat_compare_means(mapping = aes(group = Patho), label = "p.signif") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
predict(
  object = pls_res,
  newdata = select(.data = test_pre_post, -c("Patho", "Patho_num"))
) |>
  as.factor() |>
  confusionMatrix(data = test_pre_post$Patho |> as.factor()) |>
  draw_confusion_matrix()
validationplot(object = pls_res, val.type = "RMSE")
validationplot(object = pls_res, val.type = "R2")
```

# Random comparison

We compare data from pathological subjects with data from non pathological ones on the same day to avoid differences induced by the hypoxic exposure.

## k-means

```{r}
#| label: km-rand
#| fig-align: center
km_res <- kmeans(
  x = select(.data = data_rand, -c("Patho", "Patho_num")),
  centers = 2L,
  nstart = 25L
)
km_cluster <- km_res$cluster - 1
fviz_cluster(
  object = km_res,
  data = select(.data = data_rand, -c("Patho", "Patho_num")),
  geom = "none"
) +
  geom_point(mapping = aes(shape = data_rand$Patho), size = 2) +
  labs(shape = "Status")
confusionMatrix(data = as.factor(x = km_cluster), as.factor(x = data_rand$Patho_num)) |>
  draw_confusion_matrix()
```

## Hierarchical clustering

```{r}
#| label: hc-rand
#| fig-align: center
hc_res <- agnes(x = select(.data = data_rand, -c("Patho", "Patho_num")), method = "ward")
hclust_plot(clust_data = hc_res, target = data_rand$Patho_num)
confusionMatrix(
  data = hc_res |> cutree(k = 2) |> subtract(e1 = 1) |> as.factor(),
  data_rand$Patho_num |> as.factor()
) |> draw_confusion_matrix()
```

## PCA

```{r}
#| label: pca-rand
#| fig-align: center
pca_res <- prcomp(x = select(.data = data_rand, -c("Patho", "Patho_num")))
autoplot(
  object = pca_res,
  data = select(.data = data_rand, -c("Patho", "Patho_num")),
  loadings.label.size = 3
) +
  geom_point(mapping = aes(color = data_rand$Patho), size = 2) +
  stat_ellipse(mapping = aes(group = data_rand$Patho, colour = data_rand$Patho)) +
  labs(color = "Patho")
```

## PLS-DA

```{r}
#| label: plsda-rand
#| fig-align: center
pls_res <- mixOmics::plsda(
  X = select(.data = data_rand, -c("Patho", "Patho_num")),
  Y = as.factor(x = data_rand$Patho),
  ncomp = 5
)
pc1_var <- pls_res$prop_expl_var$X[[1]] * 100
pc2_var <- pls_res$prop_expl_var$X[[2]] * 100
ggplot(
  data = pls_res$input.X,
  mapping = aes(x = pls_res$variates$X[, 1], y = pls_res$variates$X[, 2])
) +
  geom_point(mapping = aes(color = data_rand$Patho), size = 2) +
  stat_ellipse(mapping = aes(group = data_rand$Patho, colour = data_rand$Patho)) +
  xlab(label = paste0("PC1 (", pc1_var |> round(digits = 1), "%)")) +
  ylab(label = paste0("PC2 (", pc2_var |> round(digits = 1), "%)")) +
  labs(color = "Patho")
pls_res <- caret::plsda(
  x =
    select(.data = data_rand, -c("Patho", "Patho_num")),
  as.factor(x = data_rand$Patho),
  ncomp = 5,
  validation = "CV"
)
my_vips <- vip(object = pls_res, metric = "rmse")
plot(x = my_vips)
plot_data <- data_rand |>
  select(c("Patho", my_vips$data$Variable)) |>
  pivot_longer(
    cols = starts_with(match = "B"),
    names_to = "Buckets",
    values_to = "Value"
  )
buckets_stats <- lapply(
  my_vips$data$Variable,
  \(x) wilcox.test(data_rand[data_rand$Patho == "Pathological", x], data_rand[data_rand$Patho == "Healthy", x])$p.value
) |>
  `names<-`(my_vips$data$Variable) |>
  lapply(p.adjust, method = "BH")
ggplot(
  data = plot_data,
  mapping = aes(x = Buckets, y = Value, fill = Patho)
) +
  geom_boxplot() +
  stat_compare_means(mapping = aes(group = Patho), label = "p.signif") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
predict(object = pls_res, newdata = select(.data = test_rand, -c("Patho", "Patho_num"))) |>
  as.factor() |>
  confusionMatrix(data = test_rand$Patho |> as.factor()) |>
  draw_confusion_matrix()
validationplot(object = pls_res, val.type = "RMSE")
validationplot(object = pls_res, val.type = "R2")
```
